```{r, echo = FALSE}
title <- "deconvolution"
```

---
title: `r title`
author: "Steffan Noe Niikanoff Christiansen"
output: html_document
---


### Global options and paths
```{r Global options and paths, echo = FALSE, message=FALSE}
# libraries
packages <- c("tidyverse", "patchwork", "noecodedup", "here", "knitr", "EpiSCORE")

invisible(lapply(packages, require, character.only = TRUE)) 
theme_set(theme_bw() + 
            theme(plot.title = element_text(hjust = 0.5),
                  axis.text.x = element_text(angle = 45,
                                             margin = margin(4, 0, 0, 0),
                                             hjust = 1),
                  strip.text = element_text(size = 8),
                  strip.background = element_rect(fill="white"),
                  panel.spacing = unit(0.1, "lines")))
# update_geom_defaults("vline", aes(linewidth = 0.75))

# paths
root_path <- here()
projects_path <- str_remove(root_path, "/[A-z]+$")
title <- str_replace_all(title, " ", "_")
fig_path <- here("results", "images", title, "/")
tab_path <- here("results", "tables", title, "/")
ref_path <- here("data", "deconvolution", "references")

# global options
opts_knit$set(root.dir = root_path)
opts_chunk$set(message = FALSE,
               dpi = 300,
               dev = "jpeg",
               fig.width = 6,
               fig.height = 6,
               fig.align="center", 
               fig.path = fig_path,
               echo = FALSE,
               warning = FALSE)

options(scipen = 100, digits = 3, bitmapType = "cairo") # fixes grid.Call warning 
time <- lubridate::date()
```

Rendering of the report was started `r time` <br> Following packages
were loaded: `r packages` <br><br>


#### Data
```{r Data}
load("data/deconvolution/SkinRef.rda")
load("data/deconvolution/VandiverDNAm.rda")
```
*Upon loading this data object, there are two reference matrices, one for gene-expression (Skin_Expref.m) and another for DNA methylation (Skin_Mref.m). The DNAm reference matrix for skin is defined for 111 marker genes and 7 cell-types which include endothelial cells (EC), fibroblasts (Fib), differentiated and undifferentiated keratinocytes (Kera_diff & Kera_undiff), macrophages (Macro), melanocytes (Mela) and T-cells (Tcells).*
<br><br>

`Skin_Expref.m`
```{r , fig.width = 10, fig.height = 6, out.width = "100%"}
Skin_Expref.m %>% head
```
<br><br>

`Skin_Mref.m`
```{r name, fig.width = 10, fig.height = 6, out.width = "100%"}
Skin_Mref.m %>% head
```
<br><br>


*To validate the DNAm reference matrix, we can apply it to an Illumina 450k DNAm skin dataset with samples from dermis and epidermis, since the epidermis contains mostly keratinocytes whilst the dermis is mostly composed of fibroblasts and endothelial cells. A version of this dataset where DNAm is summarized at the gene-level (NCBI EntrezIDs), is provided, so we first load it*

```{r, echo = TRUE}
VandiverDNAm.m %>% dim
VandiverDNAm.m[1:5, 1:6]
```
Notice: <br>

  * 12117 rows of object. Range of values: `r range(VandiverDNAm.m)`
<br><br>


#### Cell-type estimation
*The next step is to then estimate cell-type fractions in this data, which can be accomplished using the wRPC function of the EpiSCORE R-package:*
```{r, echo = TRUE}
estF.o <- wRPC(
  VandiverDNAm.m,
  ref=Skin_Mref.m,
  useW=TRUE,
  wth=0.4,
  maxit=200
)
```

* In the above, the parameter useW=TRUE, means that we are going to use the weights (a quality/confidence score for each marker gene) when performing the multivariate regression, and the threshold wth=0.4, specifies that we will only use marker genes with weights larger than 0.4 when performing the weighted regression. The last parameter maxit  specifies the maximum number of iterations to perform. 
The estimated cell-type fraction matrix, i.e. a matrix with rows labeling the samples and columns labeling the 7 cell-types can be found in the output object as `estF.o$est`. Finally, one would check that the fractions for endothelial and fibroblast cells are highest in the dermis, whilst the fractions for the keratinocytes are highest in the epidermis. Meta-information regarding the skin samples, including dermis/epidermis, is provided in the `PhenoTypesVAND.lv$Tissue object`.*

```{r Cell-type estimation, fig.width = 10, fig.height = 6, out.width = "100%"}
estF.o$est %>% head()
```
Range of row sums in object above: `r rowSums(estF.o$est) %>% range()`
<br><br>


```{r}
shared_rownames <- intersect(rownames(Skin_Mref.m), rownames(VandiverDNAm.m))
shared_rownames
```
The rownames above (n = `r length(shared_rownames)`) are shared among the reference matrix and the data. 




### References
Downloaded from https://www.biosino.org/episcore/
```{r References, fig.width = 10, fig.height = 6, out.width = "100%"}
biosino_skin_ref <- read_csv(here(ref_path, "Skin_Mref.csv"))

biosino_skin_ref[1:3,]
Skin_Mref.m[1:3,]
```
The object downloaded from biosino (`biosino_skin_ref`) and the object provided in supplementary material (`Skin_Mref.m`) seems to be identical. 
<br>

#### Heart reference
```{r}
biosino_heart_ref <- read_csv(here(ref_path, "Heart_Mref.csv"))
biosino_heart_ref

biosino_heart_ref_mat <- as.matrix(biosino_heart_ref)
rownames(biosino_heart_ref_mat) <- biosino_heart_ref_mat[,1]
biosino_heart_ref_mat <- biosino_heart_ref_mat[, -1] # removing gene_names before cell type estimation
```
Notice that five different cell types are estimated. The density of weights are shown below:

```{r Heart references, fig.width = 10, fig.height = 6, out.width = "80%"}
p1 <- biosino_heart_ref %>% 
  ggplot(aes(x = weight)) + 
  geom_density() + 
  geom_vline(xintercept = 0.4, color = "red") +
  ggtitle("Weights of 207 genes")

p2 <- biosino_heart_ref %>% 
  # filter(weight > 0.4) %>% 
  pivot_longer(cols = CM:SMC) %>% 
  ggplot(aes(x = value, color = name)) + 
  geom_density() + 
  geom_vline(xintercept = 0.4, color = "red") +
  ggtitle("Values of 207 genes colored by cell type")




# p2 <- biosino_heart_ref %>% 
#   filter(weight > 0.4) %>% 
#   pivot_longer(cols = CM:SMC) %>% 
#   ggplot(aes(x = name, y = value)) + 
#   geom_point(alpha = 0.25) +
#   geom_boxplot() + 
#   geom_hline(yintercept = 0.4, color = "red") +
#   ggtitle("112 genes with weight >0.4")
  
p1 + p2 + plot_layout(ncol = 2)
```
Notice: <br>

  * According to tutorial, weights above 0.4 should be used for deconvolution. Among the `r length(biosino_heart_ref$weight)` genes,  
`r length(which(biosino_heart_ref$weight>0.4))` passes this threshold. 
  * Seems values (not weights) are lower for CM, EC, and FB. Affecting end results?

```{r}
beta_values <- readRDS("results/processed/beta_values_combat_2024_06_27.rds")
sample_sheet <- readRDS("data/sample_sheet_2024_10_23.rds")
selected_colors <- readRDS("data/selected_colors_2023_12_19.rds")
sample_sheet <- sample_sheet %>% 
  mutate(id = factor(id))

promotor_data_matrix_EPICv2 <- constAvBetaTSS(
  beta.m = beta_values,
  type = "EPICv2")

# promotor_data_matrix_450k <- constAvBetaTSS(
#   beta.m = beta_values,
#   type = "450k")
# promotor_data_matrix_all <- constAvBetaTSS(
#   beta.m = beta_values,
#   type = "850k")

# dim(promotor_data_matrix_EPICv2)
```
Promoter averages is calculated using the `constAvBetaTSS` function. The function is described in section 2.5 of the vignette (`vignette("EpiSCORE")`). <br>

### Ventricle cell type estimation
```{r, fig.width = 10, fig.height = 6, out.width = "100%"}
est_ventricles <- wRPC(
  promotor_data_matrix_EPICv2,
  ref=biosino_heart_ref_mat,
  useW=TRUE,
  wth=0.4,
  maxit=200 # test
)

# est_maxit_50 <- wRPC(
#   promotor_data_matrix_EPICv2,
#   ref=biosino_heart_ref_mat,
#   useW=TRUE,
#   wth=0.4,
#   maxit=50)

est_maxit_200 <- wRPC(
  promotor_data_matrix_EPICv2,
  ref=biosino_heart_ref_mat,
  useW=TRUE,
  wth=0.4,
  maxit=200)
# 
# est_maxit_800 <- wRPC(
#   promotor_data_matrix_EPICv2,
#   ref=biosino_heart_ref_mat,
#   useW=TRUE,
#   wth=0.4,
#   maxit=800)

# est_maxit_50$estF[1:3,]
est_maxit_200$estF[1:3,]
# est_maxit_800$estF[1:3,]

# change in paramteres yield close to no difference at all


est_ventricles_df <- data.frame(est_ventricles$estF) %>% 
  rownames_to_column(., var = "id_pos") %>% 
  left_join(., sample_sheet)
```


```{r}
est_ventricles_df %>% 
  select(CM:SMC, id, origin) %>% 
  saveRDS(., "results/processed/cell_types/est_ventricles_df.rds")
```


```{r Ventricle cell type estimation, fig.width = 10, fig.height = 6, out.width = "100%"}
est_ventricles_df %>% 
  select(id, origin, CM:SMC) %>% 
  pivot_longer(cols = CM:SMC, 
               names_to = "type", 
               values_to = "fraction") %>% 
  ggplot(aes(x = id, y = fraction, fill = type)) +
  geom_bar(stat = "identity") +
  facet_wrap(~origin)

samp_fraction_above_002 <- est_ventricles_df %>% 
  filter(CM > 0.02)
```
Notice: <br>

  * `$ref` of the `wRPC` output holds the same information reference used as input
  * `r nrow(samp_fraction_above_002)` samples had a CM fraction above 0.02 (2%!). Max CM fraction is `r max(est_ventricles_df$CM) `
<br><br>

The number of markers per cell-type is stated below:
```{r , fig.width = 10, fig.height = 6, out.width = "100%"}
est_ventricles$ref %>% 
  as.data.frame() %>% 
  pivot_longer(cols = CM:SMC) %>% 
  count(name, value > 0.4) %>% 
  filter(`value > 0.4`) %>% 
  s_table()
```
<br><br>

Median per tissue and cell type is shown below:
```{r}
est_ventricles_df %>% 
  group_by(origin) %>% 
  reframe(median_CM = median(CM),
          median_SMC = median(SMC),
          median_EC = median(EC),
          median_FB = median(FB),
          median_MP = median(MP)) %>% 
  s_table()
```
<br><br>


#### Collapsed cell type estimation
Based on observation, a [github request](https://github.com/aet21/EpiSCORE/issues/18) was opened. Developed of EpiSCORE states following: *The relative proportion of cardiomyocytes (CM) and smooth muscle cells (SMC) is know to vary very substantially depending on where the samples derive from. If from the aorta, the CM fraction should indeed be very low, as SMC is the dominant fraction and I note that in our publication we validated the Heart DNAm reference against aortic dissection and BAV samples, where the CM fraction was indeed very low. I presume therefore that your samples are from the myocardium, as you expect much higher CM fractions. From what you state, it would appear that our Heart DNAm panel is underestimating the true CM fraction by around 20-30%, and my instinct would be the estimated SMC fraction will correspondingly be overestimated by that amount, although you don't state that. Because both are muscle cell-types, it is possible that there is some confusion between the two if the marker genes that we identified from scRNA-Seq data are not distinguished via promoter DNAm levels. This is certainly possible, or alternatively, the lower number of CMs in the scRNA-Seq assay may have compromised the detecting of CM-specific markers.* <br><br>

To fix the issue, a collapsed analysis was conducted based on SMC + CM. 
```{r Collapsed cell type estimation, fig.width = 10, fig.height = 6, out.width = "100%"}
est_ventricles_collapsed <- est_ventricles_df %>% 
  as.data.frame() %>%
  mutate(CM_SMC = CM + SMC) %>% 
  select(id, origin, EC, FB, MP, CM_SMC) %>%
  pivot_longer(cols = EC:CM_SMC, 
               names_to = "type", 
               values_to = "fraction")

est_ventricles_collapsed %>% 
  ggplot(aes(x = id, y = fraction, fill = type)) +
  geom_bar(stat = "identity") +
  facet_wrap(~origin)

```
Notice: <br>

  * The FB value ranges from `r min(est_ventricles_df$FB)` to `r max(est_ventricles_df$FB)` meaning that CM to FB ratios are meaningless (Inf values produced). 
  
#### Collapsed grouped by ID
```{r Collapsed grouped by ID, fig.width = 10, fig.height = 6, out.width = "100%"}
est_ventricles_collapsed %>% 
  ggplot(aes(x = origin, y = fraction, fill = type)) +
    geom_bar(stat = "identity") +
    facet_wrap(~id, ncol = 3)
```
<br><br>
  
<br><br>



#### Unresolved questions
List: <br> 

  * `We can see that the estimated keratinocyte fraction is high in epidermis and low in dermis`- check
  * CM estimates close to expected?

 <br><br> 


<br><br>

------------------------------------------------------------------------
```{r computetime, echo=FALSE}
time <- lubridate::date()
```
Report was finished `r time` <br>


```{r}
sessionInfo()
```


```{r rendering while keeping objects in global environment, echo = FALSE, eval=FALSE}
rmarkdown::render("scripts/deconvolution.Rmd", clean = TRUE, output_dir = "reports/")
```
<br>
